100 years of Pies vs.~Bars

### 2024-12-03 index.qmd
Notes Integrated

William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions had been publicized in textbooks and manuals, and statisticians began conducting experimental studies comparing the two.]{.svp}
[These studies have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.\]{.svp}

[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
[We begin with an initial set of four studies from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results. Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}
[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}

@@

# Summary from Olivia's note

Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships. However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams. Accuracy was influenced by the arrangement of segments (e.g., circles with sections centered at specific angles were more straightforward to interpret). Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons. They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes. Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings. These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.
Overall, the studies collectively demonstrate that the choice of visualization depends on the task and the data structure. Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components. Visual aids, such as scales and labels, further enhance accuracy and usability.

# Summary from Mason's note

Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability. Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated. Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures. Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40–250% higher errors. Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors. Biases and outliers were common, with position accuracy declining as distances increased. These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension. These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.

>>>>

William Playfair introduced the pie and bar charts in the early 1800s[, and by the turn of the next century, strong opinions had been publicized in textbooks and manuals, and statisticians began conducting experimental studies comparing the two.]{.svp}
[These studies have contradictory results:]{.svp} @vonhuhnFurtherStudiesGraphic1927 concludes that bar charts are much more useful and efficient in the majority of situations, even though pie charts have specific applications[, while @croxtonBarChartsCircle1927 found that pie charts may match or surpass them in certain situations.]{.svp}
[Much later, @clevelandGraphicalPerceptionTheory1984 conducted experiments on simple graphical elements, stating that aligned length judgments are more accurate than judgments based on area or angle, with the implication that bar charts are more accurately perceived than pie charts.]{.svp}
[Their hierarchy of feature comprehension is often assumed to be experimentally derived, but the experiments presented in the papers [@clevelandGraphicalPerceptionTheory1984; @clevelandGraphicalPerceptionGraphical1985; @clevelandExperimentGraphicalPerception1986] only covered a few elements of this hierarchy.]{.svp}

[Here, we re-examine the historical literature surrounding the use of pie and bar charts, examining recommendations about the use of pie and bar charts, tracing these guidelines back to empirical studies, and evaluating whether the recommendations are well-supported by the studies.]{.svp}
[We begin with an initial set of four studies from the late 1920s and early 1930s: @eellsRelativeMeritsCircles1926, @croxtonBarChartsCircle1927, @vonhuhnFurtherStudiesGraphic1927, and @croxtonGraphicComparisonsBars1932.]{.svp}
[We trace these studies forward in time, examining the claims made about the studies as well as the recommendations made in citing papers.]{.svp}
[In addition, we lay out the different experimental designs used in each experimental study, with the goal of assessing how the design of each study might contribute to the ultimate results.]{.svp}
[They demonstrated that the choice of visualization depends on the task and the data structure.
Bars are preferable for comparisons and time-series data, while circles may be suitable for single distributions with grouped components.]{.mt}
[Eells (1926, 1927) found that pie charts (circles) were more accurate than bar charts when presenting simple distributions, such as 50-50 or 25-75 relationships. However, bar charts were superior for more complex distributions and comparative tasks due to their linear scale, which facilitates comparisons across multiple diagrams.]{.mt}
[Vohn (1927) argued that bar charts offer advantages for tasks requiring comparisons across time series or multiple distributions, owing to their standard scale and ease of labeling, while circles are more suited for showing numerous components grouped into categories.
Frederic and Croxton (1932) compared bars, squares, circles, and cubes for simple comparisons.
They confirmed that bar charts yielded the most accurate estimates, outperforming circles, squares, and cubes.
Squares and circles were comparable in accuracy, while cubes consistently performed the worst due to the challenge of representing three-dimensional volume in two-dimensional drawings.
These findings emphasize that linear encodings (e.g., bars) are more effective than area- or volume-based encodings for precise quantitative judgments.]{.mt} [Asking for absolute estimates of differences (A - B) likely gives the advantage to bar charts, while asking for part-to-whole estimates (A/(A+B)) would be expected to give the advantage to pie charts; asking for A/B ratio comparisons may produce yet a different result.]{.svp}

[In late 70's and 80's Studies on graphical perception reveal that position is the most accurate visual encoding method, followed by length and angle, with tasks involving area, volume, curvature, shading, and color saturation needing more reliability.
Baird (1970) demonstrated that length is perceived more accurately than area and volume, which are often underestimated.
Baird and Noma (1978) highlighted how visual cues, such as frames, improve accuracy by providing reference structures.
Experiments comparing position and length judgments found that position tasks were significantly more accurate, with length judgments resulting in 40--250% higher errors.
Similarly, comparisons of position and angle judgments revealed that pie charts, which rely on angle encodings, are far less effective than bar charts, with position judgments being nearly twice as accurate and accounting for significantly fewer errors.
Biases and outliers were common, with position accuracy declining as distances increased.
These findings emphasize the need to avoid traditional visualizations like pie and divided bar charts, recommending alternatives such as dot charts or framed rectangular charts, which leverage position-based encodings to improve accuracy and comprehension.
These insights underscore the importance of redesigning data visualizations to optimize perceptual accuracy and usability.]{.mt}

[While these studies are the earliest systematic studies of statistical graphics we could locate, modern statistical graphics research might be said to date from Cleveland & McGill's work on the perception of simple graphical elements; to include this work as well, we also look forward and backward from these papers, examining both cited papers and papers describing their work in subsequent research.]{.svp}

@@

### 2024-12-03 index.tex
Notes Integrated

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

William Playfair introduced the pie and bar charts in the early
1800s\svp{, and by the turn of the next century, strong opinions had
been publicized in textbooks and manuals, and statisticians began
conducting experimental studies comparing the two.} \svp{These studies
have contradictory results:} Huhn (1927) concludes that bar charts are
much more useful and efficient in the majority of situations, even
though pie charts have specific applications\svp{, while Croxton and
Stryker (1927) found that pie charts may match or surpass them in
certain situations.} \svp{Much later, Cleveland and McGill (1984)
conducted experiments on simple graphical elements, stating that aligned
length judgments are more accurate than judgments based on area or
angle, with the implication that bar charts are more accurately
perceived than pie charts.} Their hierarchy of feature comprehension is
often assumed to be experimentally derived, but the experiments
presented in the papers (Cleveland and McGill 1984, 1985, 1986) only
covered a few elements of this hierarchy.{]}\{.svp\}

\svp{Here, we re-examine the historical literature surrounding the use
of pie and bar charts, examining recommendations about the use of pie
and bar charts, tracing these guidelines back to empirical studies, and
evaluating whether the recommendations are well-supported by the
studies.} \svp{We begin with an initial set of four studies from the
late 1920s and early 1930s: Eells (1926), Croxton and Stryker (1927),
Huhn (1927), and Croxton and Stein (1932).} \svp{We trace these studies
forward in time, examining the claims made about the studies as well as
the recommendations made in citing papers.} \svp{In addition, we lay out
the different experimental designs used in each experimental study, with
the goal of assessing how the design of each study might contribute to
the ultimate results. Asking for absolute estimates of differences (A -
B) likely gives the advantage to bar charts, while asking for
part-to-whole estimates (A/(A+B)) would be expected to give the
advantage to pie charts; asking for A/B ratio comparisons may produce
yet a different result.} \svp{While these studies are the earliest
systematic studies of statistical graphics we could locate, modern
statistical graphics research might be said to date from Cleveland \&
McGill's work on the perception of simple graphical elements; to include
this work as well, we also look forward and backward from these papers,
examining both cited papers and papers describing their work in
subsequent research.}

@@

\hypertarget{summary-from-olivias-note}{%
\section{Summary from Olivia's note}\label{summary-from-olivias-note}}

Eells (1926, 1927) found that pie charts (circles) were more accurate
than bar charts when presenting simple distributions, such as 50-50 or
25-75 relationships. However, bar charts were superior for more complex
distributions and comparative tasks due to their linear scale, which
facilitates comparisons across multiple diagrams. Accuracy was
influenced by the arrangement of segments (e.g., circles with sections
centered at specific angles were more straightforward to interpret).
Vohn (1927) argued that bar charts offer advantages for tasks requiring
comparisons across time series or multiple distributions, owing to their
standard scale and ease of labeling, while circles are more suited for
showing numerous components grouped into categories. Frederic and
Croxton (1932) compared bars, squares, circles, and cubes for simple
comparisons. They confirmed that bar charts yielded the most accurate
estimates, outperforming circles, squares, and cubes. Squares and
circles were comparable in accuracy, while cubes consistently performed
the worst due to the challenge of representing three-dimensional volume
in two-dimensional drawings. These findings emphasize that linear
encodings (e.g., bars) are more effective than area- or volume-based
encodings for precise quantitative judgments. Overall, the studies
collectively demonstrate that the choice of visualization depends on the
task and the data structure. Bars are preferable for comparisons and
time-series data, while circles may be suitable for single distributions
with grouped components. Visual aids, such as scales and labels, further
enhance accuracy and usability.

\hypertarget{summary-from-masons-note}{%
\section{Summary from Mason's note}\label{summary-from-masons-note}}

Studies on graphical perception reveal that position is the most
accurate visual encoding method, followed by length and angle, with
tasks involving area, volume, curvature, shading, and color saturation
needing more reliability. Baird (1970) demonstrated that length is
perceived more accurately than area and volume, which are often
underestimated. Baird and Noma (1978) highlighted how visual cues, such
as frames, improve accuracy by providing reference structures.
Experiments comparing position and length judgments found that position
tasks were significantly more accurate, with length judgments resulting
in 40--250\% higher errors. Similarly, comparisons of position and angle
judgments revealed that pie charts, which rely on angle encodings, are
far less effective than bar charts, with position judgments being nearly
twice as accurate and accounting for significantly fewer errors. Biases
and outliers were common, with position accuracy declining as distances
increased. These findings emphasize the need to avoid traditional
visualizations like pie and divided bar charts, recommending
alternatives such as dot charts or framed rectangular charts, which
leverage position-based encodings to improve accuracy and comprehension.
These insights underscore the importance of redesigning data
visualizations to optimize perceptual accuracy and usability.

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

William Playfair introduced the pie and bar charts in the early
1800s\svp{, and by the turn of the next century, strong opinions had
been publicized in textbooks and manuals, and statisticians began
conducting experimental studies comparing the two.} \svp{These studies
have contradictory results:} Huhn (1927) concludes that bar charts are
much more useful and efficient in the majority of situations, even
though pie charts have specific applications\svp{, while Croxton and
Stryker (1927) found that pie charts may match or surpass them in
certain situations.} \svp{Much later, Cleveland and McGill (1984)
conducted experiments on simple graphical elements, stating that aligned
length judgments are more accurate than judgments based on area or
angle, with the implication that bar charts are more accurately
perceived than pie charts.} \svp{Their hierarchy of feature
comprehension is often assumed to be experimentally derived, but the
experiments presented in the papers (Cleveland and McGill 1984, 1985,
1986) only covered a few elements of this hierarchy.}

\svp{Here, we re-examine the historical literature surrounding the use
of pie and bar charts, examining recommendations about the use of pie
and bar charts, tracing these guidelines back to empirical studies, and
evaluating whether the recommendations are well-supported by the
studies.} \svp{We begin with an initial set of four studies from the
late 1920s and early 1930s: Eells (1926), Croxton and Stryker (1927),
Huhn (1927), and Croxton and Stein (1932).} \svp{We trace these studies
forward in time, examining the claims made about the studies as well as
the recommendations made in citing papers.} \svp{In addition, we lay out
the different experimental designs used in each experimental study, with
the goal of assessing how the design of each study might contribute to
the ultimate results.} \mt{They demonstrated that the choice of
visualization depends on the task and the data structure. Bars are
preferable for comparisons and time-series data, while circles may be
suitable for single distributions with grouped components.} \mt{Eells
(1926, 1927) found that pie charts (circles) were more accurate than bar
charts when presenting simple distributions, such as 50-50 or 25-75
relationships. However, bar charts were superior for more complex
distributions and comparative tasks due to their linear scale, which
facilitates comparisons across multiple diagrams.} \mt{Vohn (1927)
argued that bar charts offer advantages for tasks requiring comparisons
across time series or multiple distributions, owing to their standard
scale and ease of labeling, while circles are more suited for showing
numerous components grouped into categories. Frederic and Croxton (1932)
compared bars, squares, circles, and cubes for simple comparisons. They
confirmed that bar charts yielded the most accurate estimates,
outperforming circles, squares, and cubes. Squares and circles were
comparable in accuracy, while cubes consistently performed the worst due
to the challenge of representing three-dimensional volume in
two-dimensional drawings. These findings emphasize that linear encodings
(e.g., bars) are more effective than area- or volume-based encodings for
precise quantitative judgments.} \svp{Asking for absolute estimates of
differences (A - B) likely gives the advantage to bar charts, while
asking for part-to-whole estimates (A/(A+B)) would be expected to give
the advantage to pie charts; asking for A/B ratio comparisons may
produce yet a different result.}

\mt{In late 70's and 80's Studies on graphical perception reveal that
position is the most accurate visual encoding method, followed by length
and angle, with tasks involving area, volume, curvature, shading, and
color saturation needing more reliability. Baird (1970) demonstrated
that length is perceived more accurately than area and volume, which are
often underestimated. Baird and Noma (1978) highlighted how visual cues,
such as frames, improve accuracy by providing reference structures.
Experiments comparing position and length judgments found that position
tasks were significantly more accurate, with length judgments resulting
in 40--250\% higher errors. Similarly, comparisons of position and angle
judgments revealed that pie charts, which rely on angle encodings, are
far less effective than bar charts, with position judgments being nearly
twice as accurate and accounting for significantly fewer errors. Biases
and outliers were common, with position accuracy declining as distances
increased. These findings emphasize the need to avoid traditional
visualizations like pie and divided bar charts, recommending
alternatives such as dot charts or framed rectangular charts, which
leverage position-based encodings to improve accuracy and comprehension.
These insights underscore the importance of redesigning data
visualizations to optimize perceptual accuracy and usability.}

\svp{While these studies are the earliest systematic studies of
statistical graphics we could locate, modern statistical graphics
research might be said to date from Cleveland \& McGill's work on the
perception of simple graphical elements; to include this work as well,
we also look forward and backward from these papers, examining both
cited papers and papers describing their work in subsequent research.}

@@
